---
title: "Case Studies"
output:
  html_document:
    theme: flatly
    css: custom.css
    highlight: null
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 8.5, fig.height = 5.5, dpi = 150)
library(tidyverse)
library(knitr)

canonical_df <- "data/df_pct_with_predictions.rds"
if (!file.exists(canonical_df)) {
  stop("Canonical predictions not found. Run /Users/samberry/Desktop/Capstone/update_data_new.R first.")
}
df_pct <- readRDS(canonical_df)

source("plot-fonts.R")

county_row <- function(cname, sname) {
  df_pct %>% filter(county == cname, state_name == sname) %>% slice(1)
}

r_jasper     <- county_row("Jasper",      "Illinois")
r_martin     <- county_row("Martin",      "Indiana")
r_rockisland <- county_row("Rock Island", "Illinois")
r_rusk       <- county_row("Rusk",        "Wisconsin")
r_lawrence   <- county_row("Lawrence",    "Illinois")
r_gogebic    <- county_row("Gogebic",     "Michigan")
```

# Six Counties, Three Stories

The six case-study counties were drawn from Model 4 residuals — the comprehensive specification — and represent three distinct patterns relative to model predictions. **Over-performers** suggest something beyond demographics is mobilizing voters: institutional access, civic infrastructure, or political context. **As-expected** counties serve as baselines where structural conditions straightforwardly predict participation. **Under-performers** raise questions about suppression, disengagement, or barriers not captured by the model.

In each tail, one county represents the most extreme case and a second the more moderate outlier — letting the analysis show the pattern holds beyond the furthest tails. The two as-expected counties sit adjacent to the median rank with residuals near zero.

---

```{r dumbbell, fig.height=3.8, fig.cap="Actual vs. predicted turnout for all six case-study counties. Colored dot = observed turnout; gray dot = Model 4 prediction. Gap labels show the residual in percentage points."}

tier_colors <- c(
  "Over-performer"  = "#D85A30",
  "As-expected"     = "#888780",
  "Under-performer" = "#7F77DD"
)

case_data <- tibble(
  county_label = c("Jasper, IL", "Martin, IN",
                   "Rock Island, IL", "Rusk, WI",
                   "Lawrence, IL", "Gogebic, MI"),
  tier = c("Over-performer", "Over-performer",
           "As-expected",    "As-expected",
           "Under-performer","Under-performer"),
  actual = c(
    r_jasper$turnout_rate,    r_martin$turnout_rate,
    r_rockisland$turnout_rate, r_rusk$turnout_rate,
    r_lawrence$turnout_rate,  r_gogebic$turnout_rate
  ) * 100,
  predicted = c(
    r_jasper$predicted4,    r_martin$predicted4,
    r_rockisland$predicted4, r_rusk$predicted4,
    r_lawrence$predicted4,  r_gogebic$predicted4
  ) * 100
) %>%
  mutate(
    county_label = factor(county_label, levels = c(
      "Gogebic, MI", "Lawrence, IL",
      "Rusk, WI", "Rock Island, IL",
      "Martin, IN", "Jasper, IL"
    )),
    tier = factor(tier, levels = c("Over-performer", "As-expected", "Under-performer")),
    gap  = actual - predicted,
    gap_label = ifelse(gap >= 0,
                       paste0("+", round(gap, 1), " pp"),
                       paste0(round(gap, 1), " pp")),
    label_hjust = ifelse(gap >= 0, -0.40, 1.40)
  )

ggplot(case_data, aes(y = county_label)) +
  geom_segment(
    aes(x    = pmin(actual, predicted),
        xend = pmax(actual, predicted),
        yend = county_label,
        color = tier),
    linewidth = 1.5, alpha = 0.40
  ) +
  geom_point(aes(x = predicted),
             color = "#C9C5BE", size = 5.5) +
  geom_point(aes(x = actual, color = tier),
             size = 6.5) +
  geom_text(
    aes(x = actual, label = gap_label,
        color = tier, hjust = label_hjust),
    family = plot_sans, size = 3.0, fontface = "bold"
  ) +
  scale_color_manual(values = tier_colors, guide = "none") +
  facet_grid(tier ~ ., scales = "free_y", space = "free_y",
             labeller = labeller(tier = c(
               "Over-performer"  = "Over-\nperformer",
               "As-expected"     = "As-\nexpected",
               "Under-performer" = "Under-\nperformer"
             ))) +
  scale_x_continuous(
    labels = scales::label_number(suffix = "%"),
    expand = expansion(mult = c(0.14, 0.16))
  ) +
  labs(
    title    = "Actual vs. Predicted Turnout — Six Case Study Counties",
    subtitle = "Colored dot = actual turnout  ·  Gray dot = Model 4 prediction",
    x = "Turnout rate (%)", y = NULL
  ) +
  theme_minimal(base_family = plot_sans, base_size = 13) +
  theme(
    plot.title    = element_text(family = plot_serif, face = "bold",
                                 size = 18, color = "#1F1F1F",
                                 margin = margin(b = 4)),
    plot.subtitle = element_text(family = plot_serif, face = "italic",
                                 size = 12, color = "#7A746C",
                                 margin = margin(b = 14)),
    strip.text.y  = element_text(family = plot_sans, face = "bold",
                                 size = 8.5, color = "#4F4A44",
                                 angle = 0, hjust = 0,
                                 lineheight = 1.15),
    strip.background = element_rect(fill = "#EEEAE4", color = NA),
    panel.grid.major.x = element_line(color = "#F1EDE6", linewidth = 0.4),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.spacing      = unit(1, "lines"),
    axis.text.y  = element_text(family = plot_sans, size = 11.5,
                                 color = "#4F4A44"),
    axis.text.x  = element_text(family = plot_sans, size = 10,
                                 color = "#7A746C"),
    axis.title.x = element_text(family = plot_sans, size = 10.5,
                                 color = "#5A544C", margin = margin(t = 8)),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(12, 70, 14, 14)
  )
```

---

## County Profiles

<style>
.county-expand-table {
  width: 100%;
  border-collapse: collapse;
  font-family: 'IBM Plex Sans', Arial, sans-serif;
  font-size: 14px;
  margin: 1rem 0 2.2rem 0;
}
.county-expand-table thead tr {
  background: #EEEAE4;
  color: #5A544C;
  font-size: 14px;
  font-weight: 600;
  text-transform: uppercase;
  letter-spacing: 0.055em;
}
.county-expand-table th {
  padding: 0.75rem 1.2rem;
  font-weight: 600;
  border-bottom: 2px solid #DDD8D0;
  white-space: nowrap;
}
.county-expand-table td {
  padding: 1.1rem 1.2rem;
  border-bottom: 1px solid #EDE9E3;
  vertical-align: middle;
}
.tier-header td {
  font-weight: 600;
  font-size: 13px;
  letter-spacing: 0.09em;
  text-transform: uppercase;
  padding: 0.45rem 1.1rem;
  border-bottom: 1px solid #E5E0D8;
}
.tier-header.over td   { background: #F5C4B3 !important; color: #993C1D !important; }
.tier-header.middle td { background: #C8C6BC !important; color: #444441 !important; }
.tier-header.under td  { background: #CECBF6 !important; color: #3C3489 !important; }
.county-row {
  cursor: pointer;
  transition: background 0.14s;
}
/* per-tier row tints — target td directly since Bootstrap paints td over tr */
.county-row.over td    { background-color: #FAECE7 !important; }
.county-row.middle td  { background-color: #F1EFE8 !important; }
.county-row.under td   { background-color: #EEEDFE !important; }
.county-row.over:hover td    { background-color: #F5DDD5 !important; }
.county-row.middle:hover td  { background-color: #E8E5DC !important; }
.county-row.under:hover td   { background-color: #E0DEFC !important; }
.county-row.over.active td    { background-color: #F2D9D0 !important; }
.county-row.middle.active td  { background-color: #E4E0D7 !important; }
.county-row.under.active td   { background-color: #D8D6FA !important; }
/* expand rows: white so card sits cleanly */
.expand-row.over td   { background-color: #FFFFFF !important; }
.expand-row.middle td { background-color: #FFFFFF !important; }
.expand-row.under td  { background-color: #FFFFFF !important; }
.expand-icon {
  color: #B0AA9F;
  font-size: 0.75rem;
  width: 1.6rem;
  text-align: center;
  padding-right: 0;
}
.residual-pos { color: #D85A30; font-weight: 700; }
.residual-neg { color: #7F77DD; font-weight: 700; }
.residual-neu { color: #888780; font-weight: 600; }
.county-name-col { font-weight: 600; color: #1F1F1F; font-size: 15px; }
.key-find-col {
  font-style: italic;
  color: #4A443D;
  font-size: 14px;
  max-width: 420px;
  line-height: 1.6;
}
.expand-row { display: none; }
.expand-body {
  padding: 1.4rem 1.6rem 1.4rem 1.6rem;
  margin: 0.3rem 0.4rem 0.6rem 0;
  background: #FFFFFF;
  border: 0.5px solid #D3D1C7;
  font-size: 15px;
  line-height: 1.7;
  color: #3A3630;
}
.expand-row.over .expand-body   { border-left: 3px solid #D85A30; }
.expand-row.middle .expand-body { border-left: 3px solid #888780; }
.expand-row.under .expand-body  { border-left: 3px solid #7F77DD; }
.expand-body p { margin: 0 0 1rem 0; }
.expand-body p:last-child { margin-bottom: 0; }
</style>

<table class="county-expand-table">
<thead>
<tr>
  <th style="width:2rem;"></th>
  <th>County</th>
  <th>State</th>
  <th>Actual</th>
  <th>Predicted</th>
  <th>Gap</th>
  <th style="min-width:220px;">Key Finding</th>
</tr>
</thead>
<tbody>

<!-- ── Over-Performers ──────────────────────────────────────────────── -->
<tr class="tier-header over">
  <td colspan="7">▲&nbsp; Over-Performers</td>
</tr>

<tr class="county-row over" onclick="toggleRow('jasper')">
  <td class="expand-icon" id="icon-jasper">▶</td>
  <td class="county-name-col">Jasper</td>
  <td>IL</td>
  <td>`r round(r_jasper$turnout_rate * 100, 1)`%</td>
  <td>`r round(r_jasper$predicted4 * 100, 1)`%</td>
  <td class="residual-pos">+`r round(r_jasper$residual4 * 100, 1)` pp</td>
  <td class="key-find-col">While there is no clear differentiating factor for mobilization to account for Jasper's overperformance, the county's smaller population size, partisan cohesion, and local development coordination may help explain why turnout exceeded structural expectations.</td>
</tr>
<tr class="expand-row over" id="expand-jasper">
<td colspan="7">
<div class="expand-body">
<p>Jasper County, Illinois, has a turnout rate of 77.7% compared to a predicted rate of 66.3%, a +11.4 percentage point gap — making it the single most overperforming county in the dataset. Despite its overperformance in turnout, no formal mobilizing institution stands out. A Republican-dominated county, there were no identifiable GOTV organizations and no documented voter-mobilization initiatives leading up to the 2024 presidential election.</p>
<p>With a population of approximately 9,300, Jasper County may reflect a small, socially embedded community that the literature often links to elevated turnout norms and social pressure. In small communities, voting is socially visible: neighbors know whether their neighbors voted, religious institutions disseminate election information and apply normative pressure, and the perceived consequence of any individual vote is amplified by smaller electorates (Putnam, 2008;<sup>1</sup> Gerber, Green &amp; Larimer, 2008<sup>2</sup>).</p>
<p>Jasper also has an interesting profile economically, being in a state of relative economic momentum. While it had modest economic growth during 2024 and an increased poverty rate, inflation-adjusted GDP grew approximately 29% between 2018 and 2022, dramatically outpacing Illinois's 1% statewide growth rate over the same period. Local development organizations such as Jasper County Economic Development Inc. (JEDI), established in 2012, indicate a level of community coordination and planning in economic growth and local development. The strong Republican identification of Jasper may also contribute to higher participation in a county with a relatively cohesive local political environment.</p>
</div>
</td>
</tr>

<tr class="county-row over" onclick="toggleRow('martin')">
  <td class="expand-icon" id="icon-martin">▶</td>
  <td class="county-name-col">Martin</td>
  <td>IN</td>
  <td>`r round(r_martin$turnout_rate * 100, 1)`%</td>
  <td>`r round(r_martin$predicted4 * 100, 1)`%</td>
  <td class="residual-pos">+`r round(r_martin$residual4 * 100, 1)` pp</td>
  <td class="key-find-col">Martin's overperformance in voter turnout appears to be less about what mobilizes voters than about who lives there. NSWC Crane's federal workforce — 6,500 employees in a county of fewer than 10,000 — generates civic participation beyond what the county's demographic profile alone would predict. This case illustrates how the presence of a major federal institution and employer can drive higher turnout through compositional mechanisms as opposed to clear mobilizational efforts.</td>
</tr>
<tr class="expand-row over" id="expand-martin">
<td colspan="7">
<div class="expand-body">
<p>Martin County, Indiana has a turnout rate of 66.1% compared to a predicted rate of 57.4%, a +8.7 percentage point gap.</p>
<p>Similar to Jasper, Martin County did not have any obvious voter mobilization organizations or efforts leading up to the 2024 presidential election. While the county's overperformance does not appear to stem from formal voter outreach efforts, it may instead reflect a compositional effect tied to the local economy and workforce structure.</p>
<p>Here, the economic activity is much more pronounced than in Jasper. Martin County is home to The Naval Surface Warfare Center Crane Division (NSWC Crane), which employs roughly 6,500 federal employees and contractors, including over 3,000 STEM workers and approximately 1,000 engineers. This base has hired more than 1,300 personnel in the last three years alone. Additionally, the average wages in Martin County are approximately 6% above the national average — atypical for a rural county and reflective of Crane's concentration of high-skill federal employment. Since 2024, Martin has seen noticeable increases in both median household income and median property value — despite a stable employment rate from 2023 to 2024.</p>
<p>In a county of fewer than 10,000 residents, this large federal workforce can drastically reshape who actually lives there. The model sees Martin as a rural Indiana county, despite the fact that the workforce composition resembles a federal research community situated in a rural setting. Research on public-sector employment suggests that government employees are generally more politically active and civically engaged than the broader population — especially in professional or technical occupations like the NSWC Crane (Cooper &amp; Reggie, 2025).<sup>3</sup> The model's standard demographic predictors — education, internet access, poverty, age — do not capture the occupational civic culture that federal professional employment appears to generate independently of those characteristics</p>
<p>Local development organizations such as the Martin County Alliance reinforce this picture of community stability and coordination, and rising property values (approximately 6% in 2023–2024) signal continued economic confidence. Taken together, Crane's workforce concentration and the surrounding civic-economic infrastructure produce turnout levels that exceed what the county's rural geography and standard demographic measures would otherwise predict.</p>
</div>
</td>
</tr>

<!-- ── As-Expected ──────────────────────────────────────────────────── -->
<tr class="tier-header middle">
  <td colspan="7">◆&nbsp; As-Expected</td>
</tr>

<tr class="county-row middle" onclick="toggleRow('rockisland')">
  <td class="expand-icon" id="icon-rockisland">▶</td>
  <td class="county-name-col">Rock Island</td>
  <td>IL</td>
  <td>`r round(r_rockisland$turnout_rate * 100, 1)`%</td>
  <td>`r round(r_rockisland$predicted4 * 100, 1)`%</td>
  <td class="residual-neu">≈`r round(r_rockisland$residual4 * 100, 1)` pp</td>
  <td class="key-find-col">Rock Island serves as a baseline and validation case for the model. Its near-zero residual suggests that the model adequately captures turnout dynamics in structurally and demographically diverse urban-edge counties. This is an important result given that much of the broader dataset is composed of smaller rural counties.</td>
</tr>
<tr class="expand-row middle" id="expand-rockisland">
<td colspan="7">
<div class="expand-body">
<p>Rock Island County, Illinois, has a turnout rate of 61.1% compared to a predicted rate of 61.0% — essentially a zero residual. Here, Rock Island serves primarily as a baseline comparison case, with the central analytic question being why the model predicts turnout so accurately.</p>
<p>Part of the Quad Cities metropolitan area along the Mississippi River, Rock Island is an urban-industrial county with approximately 145,000 residents. The county is demographically diverse, with substantial Latino and Black populations, and maintains a mixed economy that includes manufacturing, logistics, healthcare, and the Rock Island Arsenal, a major federal Army installation. Unlike Martin County, however, the federal workforce does not dominate the county's economic composition, instead serving as one part of a broader economy. Historically, the county has leaned Democratic, but has remained competitive in recent cycles.</p>
<p>Rock Island does not appear to contain unusual mobilizing or suppressive dynamics. Illinois maintains a moderately accessible voting environment, and there are no clearly identifiable civic or economic features that would be expected to elevate or depress turnout beyond what the model already captures. Instead, the county reflects a relatively balanced combination of industrial, urban, racial, and institutional characteristics that align with the model's underlying predictors.</p>
<p>Overall, this makes Rock Island important precisely because it does not stand out. Despite its relatively diverse and complex demographic and economic composition, Rock Island performed almost exactly as the model predicted, suggesting the model performs well in capturing turnout dynamics in counties that are neither overwhelmingly rural nor heavily metropolitan.</p>
</div>
</td>
</tr>

<tr class="county-row middle" onclick="toggleRow('rusk')">
  <td class="expand-icon" id="icon-rusk">▶</td>
  <td class="county-name-col">Rusk</td>
  <td>WI</td>
  <td>`r round(r_rusk$turnout_rate * 100, 1)`%</td>
  <td>`r round(r_rusk$predicted4 * 100, 1)`%</td>
  <td class="residual-neu">≈`r round(r_rusk$residual4 * 100, 1)` pp</td>
  <td class="key-find-col">Rusk serves as a rural validation case. Its near-zero residual confirms that the model adequately captures turnout dynamics in typical rural counties without unusual mobilizing or suppressing factors. Paired with Rock Island, the two as-expected cases demonstrate the model's consistency across both rural and urban-edge contexts.</td>
</tr>
<tr class="expand-row middle" id="expand-rusk">
<td colspan="7">
<div class="expand-body">
<p>Rusk County, Wisconsin, has a turnout rate of 73.6% compared to a predicted rate of 73.6% — a residual of essentially zero. Like Rock Island, the analytic question for Rusk is why the model predicts it so cleanly.</p>
<p>Rusk is a small, predominantly rural county of approximately 14,000 residents in Wisconsin. It is overwhelmingly white, has lower median income levels than the state average, and is economically dependent on agriculture and light manufacturing. Wisconsin's accessible voting environment applies uniformly, and Rusk shows no unique local civic infrastructure or mobilization efforts.</p>
<p>Rusk's near-zero residual indicates that the model captures rural counties cleanly when neither unusual civic capacity nor large-scale federal employers reshape the county's composition. It is a quiet baseline case — showing how the model's structural predictions translate directly into observed turnout when the surrounding conditions are typical for the region.</p>
</div>
</td>
</tr>

<!-- ── Under-Performers ─────────────────────────────────────────────── -->
<tr class="tier-header under">
  <td colspan="7">▼&nbsp; Under-Performers</td>
</tr>

<tr class="county-row under" onclick="toggleRow('lawrence')">
  <td class="expand-icon" id="icon-lawrence">▶</td>
  <td class="county-name-col">Lawrence</td>
  <td>IL</td>
  <td>`r round(r_lawrence$turnout_rate * 100, 1)`%</td>
  <td>`r round(r_lawrence$predicted4 * 100, 1)`%</td>
  <td class="residual-neg">`r round(r_lawrence$residual4 * 100, 1)` pp</td>
  <td class="key-find-col">Lawrence's underperformance highlights the limits of structural predictors alone in explaining turnout variation. Despite several characteristics commonly associated with higher participation — including high homeownership, partisan alignment, and modest economic infrastructure — turnout falls substantially below expectations. The contrast with counties like Jasper suggests that additional local social, civic, or organizational dynamics not fully captured by the model may shape whether participation exceeds or falls below predicted levels.</td>
</tr>
<tr class="expand-row under" id="expand-lawrence">
<td colspan="7">
<div class="expand-body">
<p>Lawrence County, Illinois has a turnout rate of 50.9% compared to a predicted rate of 57.5%, a −6.6 percentage point underperformance. Despite low income and modest educational attainment, the county has a high homeownership rate and very low median property values — both of which are often associated with higher turnout. The county also operates one of only eight Foreign Trade Zones in Illinois, indicating some degree of institutional economic infrastructure.</p>
<p>Still, there is little evidence of local civic or mobilizing infrastructure that might be powering the higher participation. There were no documented GOTV organizations during the 2024 election, no prominent civic associations, and little evidence of organized voter mobilization. While Lawrence's strongly Republican political environment could provide some partisan mobilization, there is no clear organizational or institutional channel connecting the county's relatively stable population structure to higher turnout.</p>
<p>Although Lawrence shares some characteristics with counties like Jasper County — including a smaller population size, high homeownership, and a strongly Republican political environment — the county underperforms relative to expectations. The divergence between the two cases suggests that structural characteristics alone are insufficient to explain turnout variation and that unobserved local social or civic dynamics not reflected in available data or county websites may also shape voter participation in ways that are difficult to directly measure.</p>
</div>
</td>
</tr>

<tr class="county-row under" onclick="toggleRow('gogebic')">
  <td class="expand-icon" id="icon-gogebic">▶</td>
  <td class="county-name-col">Gogebic</td>
  <td>MI</td>
  <td>`r round(r_gogebic$turnout_rate * 100, 1)`%</td>
  <td>`r round(r_gogebic$predicted4 * 100, 1)`%</td>
  <td class="residual-neg">`r round(r_gogebic$residual4 * 100, 1)` pp</td>
  <td class="key-find-col">Gogebic's underperformance suggests that political realignment may outpace the rebuilding of local mobilization infrastructure. Historically, Gogebic had a strong union presence that helped structure political participation and Democratic Party engagement. As the county shifted Republican, the role of Unions appears to have weakened as well, while there is limited evidence of other equivalent institutions emerging in its place. The case highlights how turnout can fall below structural expectations during periods of organizational and political transition in formerly union-centered rural communities.</td>
</tr>
<tr class="expand-row under" id="expand-gogebic">
<td colspan="7">
<div class="expand-body">
<p>Gogebic County has a turnout rate of 70.3% compared to a predicted rate of 79.7%, a −9.4 percentage point gap.</p>
<p>What distinguishes Gogebic is its political and institutional trajectory. Historically, the county was part of a cluster of union-oriented, consistently Democratic counties in Michigan's Upper Peninsula. However, like many similar rural areas across the Midwest, Gogebic has undergone substantial political realignment, shifting from a consistently Democratic county that supported Barack Obama in 2012 to supporting Donald Trump in the past three presidential elections.</p>
<p>The county also faces considerable economic strain. Of the 7,038 households in Gogebic County in 2023, 17% earned below the Federal Poverty Level, while an additional 33% were classified as ALICE households — earning above the poverty line but still unable to consistently afford basic necessities — leaving roughly half of county households below the ALICE Threshold. Voter turnout also declined by approximately 6.5 percentage points between 2020 and 2024.</p>
<p>One possible explanation is that this political transition coincided with the weakening of older organizational structures that historically supported voter participation. While unions once played an important role in mobilizing voters, their influence appears to have weakened over time during the party realignment. While partisan preferences shifted, there is limited visible evidence that equivalent civic or organizational structures emerged to replace the motivating role unions once played. Although some nonprofit infrastructure exists within the county, there are few clearly identifiable organizations occupying a broad voter-mobilization function.</p>
<p>Gogebic's case suggests that political realignment and organizational decline can disrupt established patterns of civic participation. As partisan preferences change, the weakening of institutions that historically structured participation may contribute to county turnout falling below the model's expectations — highlighting the importance of local organizational infrastructure beyond demographic and economic characteristics alone.</p>
</div>
</td>
</tr>

</tbody>
</table>

<script>
function toggleRow(id) {
  var expRow  = document.getElementById('expand-' + id);
  var icon    = document.getElementById('icon-' + id);
  var trigger = icon.closest('tr');
  if (expRow.style.display === 'table-row') {
    expRow.style.display = 'none';
    icon.textContent = '▶';
    trigger.classList.remove('active');
  } else {
    expRow.style.display = 'table-row';
    icon.textContent = '▼';
    trigger.classList.add('active');
  }
}
</script>

---

<div class="footnotes" style="font-size:0.82rem; color:#7A746C; line-height:1.6; margin-top:1.5rem;">
<p><strong>Notes</strong></p>
<p><sup>1</sup> Putnam, R. D. (2008). <em>Bowling Alone: The Collapse and Revival of American Community.</em> Simon &amp; Schuster.</p>
<p><sup>2</sup> Gerber, A. S., Green, D. P., &amp; Larimer, C. W. (2008). Social pressure and voter turnout: Evidence from a large-scale field experiment. <em>American Political Science Review, 102</em>(1), 33–48.</p>
<p><sup>3</sup> Cooper, C. A., &amp; Reggie, J. (2025). Public sector employees' political activity across administrative traditions: Tensions between democratic participation and bureaucratic impartiality in an era of populism and democratic backsliding. <em>Review of Public Personnel Administration, 49</em>(4). https://doi.org/10.1177/07349149241310404</p>
</div>

---

<div class="site-footer">
Qualitative analysis draws on secondary sources including Michigan Department of State 2024 Election Report, Ballotpedia, NCSL voting law summaries, VoteRiders, and the Evansville NewsLab / CivicInfo Index. Visualizations produced in R with ggplot2.
</div>
