---
title: "Case Studies"
output:
  html_document:
    theme: null
    css: custom.css
    highlight: pygments
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 7.5, fig.height = 4.2, dpi = 150)
library(tidyverse)
library(tidycensus)
library(knitr)
library(kableExtra)

df_pct <- readRDS("data/df_pct.rds")

poverty_data <- get_acs(
  geography = "county",
  variables = c(poverty = "B17001_002", total = "B17001_001"),
  year = 2024, survey = "acs5"
)
poverty_pct <- poverty_data %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(pct_poverty = poverty / total * 100)

df_pct <- df_pct %>%
  left_join(poverty_pct %>% select(GEOID, pct_poverty),
            by = c("geoid" = "GEOID"))

model1 <- lm(
  turnout_rate ~
    pct_bachelors +
    pct_households_internet +
    pct_limited_english +
    pct_disability_total +
    pct_poverty,
  data = df_pct
)
df_pct$predicted <- predict(model1)
df_pct$residual  <- df_pct$turnout_rate - df_pct$predicted

# Helper: pull one county's stats cleanly
county_row <- function(cname) {
  df_pct %>% filter(county == cname) %>% slice(1)
}

# Predictor comparison helper
predictor_bar <- function(cname, state_abbr) {
  target <- df_pct %>% filter(county == cname) %>% slice(1)
  state_avg <- df_pct %>%
    filter(state_name == target$state_name) %>%
    summarise(across(c(pct_bachelors, pct_households_internet,
                       pct_limited_english, pct_disability_total, pct_poverty),
                     mean, na.rm = TRUE)) %>%
    mutate(county = paste0(target$state_name, " avg"), type = "State Avg")

  vars <- c("pct_bachelors", "pct_households_internet",
            "pct_limited_english", "pct_disability_total", "pct_poverty")
  labels <- c("Bachelor's+", "Internet Access", "Limited English",
              "Disability", "Poverty")

  bind_rows(
    target %>% select(county, all_of(vars)) %>% mutate(type = "County"),
    state_avg %>% select(county, all_of(vars), type)
  ) %>%
    pivot_longer(all_of(vars), names_to = "variable", values_to = "value") %>%
    mutate(variable = factor(variable, levels = vars, labels = labels))
}
```

<span class="section-label">Case Studies</span>

# Six Counties, Three Stories

The six case-study counties represent three distinct patterns relative to model predictions. **Over-performers** suggest that something in the civic environment — local organizing, competitive elections, cultural norms — is mobilizing voters beyond what demographics alone would predict. **Near-expected** counties serve as a baseline. **Under-performers** raise questions about suppression, disengagement, or structural barriers not captured by the model.

Each county profile includes:  
(1) key statistics from the model,  
(2) a comparison of predictors against the state average, and  
(3) space for qualitative findings from document review, news archive analysis, or other sources.

---

# Over-Performers {.tabset}

## Montmorency County, MI

```{r montmorency-stats}
r <- county_row("Montmorency")
```

<div class="county-card over" style="margin-bottom:1.5rem;">
<div class="card-label">Over-Performer · Michigan</div>
<div class="card-name">Montmorency County</div>
<div class="card-stat">Actual turnout: **`r round(r$turnout_rate, 1)`%** &nbsp;|&nbsp; Predicted: `r round(r$predicted, 1)`%</div>
<div class="card-residual">Residual: +`r round(r$residual, 2)` percentage points</div>
</div>

### Predictor Profile

```{r montmorency-bar, fig.cap="County values vs. state average for all five model predictors.", fig.height=3.5}
predictor_bar("Montmorency") %>%
  ggplot(aes(x = value, y = variable, fill = type)) +
  geom_col(position = "dodge", width = 0.65) +
  scale_fill_manual(values = c("County" = "#2e7d52", "State Avg" = "#d8d4ce")) +
  labs(x = "Percent (%)", y = NULL, fill = NULL) +
  theme_minimal(base_family = "sans", base_size = 11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA))
```

### Qualitative Analysis

> **[Placeholder — add your findings here]**

*Suggested dimensions to investigate:*

- **Civic infrastructure:** Are there active nonpartisan GOTV organizations? Strong union presence? Churches with voter mobilization history?
- **Electoral competition:** Was there a particularly competitive local race in 2020 that drove turnout above what national-level enthusiasm alone would predict?
- **Historical patterns:** Is over-performance consistent across election cycles, or is 2020 an anomaly?
- **Community characteristics:** What do residents, local news, or community organizations say about civic participation?

<div class="callout finding">
<strong>Emerging Finding</strong>
[Replace this with your key takeaway from qualitative research. What explains the gap between predicted and actual turnout?]
</div>

---

## Trempealeau County, WI

```{r trempealeau-stats}
r <- county_row("Trempealeau")
```

<div class="county-card over" style="margin-bottom:1.5rem;">
<div class="card-label">Over-Performer · Wisconsin</div>
<div class="card-name">Trempealeau County</div>
<div class="card-stat">Actual turnout: **`r round(r$turnout_rate, 1)`%** &nbsp;|&nbsp; Predicted: `r round(r$predicted, 1)`%</div>
<div class="card-residual">Residual: +`r round(r$residual, 2)` percentage points</div>
</div>

### Predictor Profile

```{r trempealeau-bar, fig.cap="County values vs. state average for all five model predictors.", fig.height=3.5}
predictor_bar("Trempealeau") %>%
  ggplot(aes(x = value, y = variable, fill = type)) +
  geom_col(position = "dodge", width = 0.65) +
  scale_fill_manual(values = c("County" = "#2e7d52", "State Avg" = "#d8d4ce")) +
  labs(x = "Percent (%)", y = NULL, fill = NULL) +
  theme_minimal(base_family = "sans", base_size = 11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA))
```

### Qualitative Analysis

> **[Placeholder — add your findings here]**

*Trempealeau County, WI — note its agricultural/rural character, proximity to La Crosse metro, and Wisconsin's historically high-turnout culture. Consider how AVR, same-day registration, and any local organizing might interact.*

<div class="callout finding">
<strong>Emerging Finding</strong>
[Replace this with your key takeaway.]
</div>

---

# Near-Expected {.tabset}

## Brown County, WI

```{r brown-stats}
r <- county_row("Brown")
```

<div class="county-card middle" style="margin-bottom:1.5rem;">
<div class="card-label">Near-Expected · Wisconsin</div>
<div class="card-name">Brown County</div>
<div class="card-stat">Actual turnout: **`r round(r$turnout_rate, 1)`%** &nbsp;|&nbsp; Predicted: `r round(r$predicted, 1)`%</div>
<div class="card-residual">Residual: `r round(r$residual, 2)` percentage points</div>
</div>

### Predictor Profile

```{r brown-bar, fig.cap="County values vs. state average.", fig.height=3.5}
predictor_bar("Brown") %>%
  ggplot(aes(x = value, y = variable, fill = type)) +
  geom_col(position = "dodge", width = 0.65) +
  scale_fill_manual(values = c("County" = "#888580", "State Avg" = "#d8d4ce")) +
  labs(x = "Percent (%)", y = NULL, fill = NULL) +
  theme_minimal(base_family = "sans", base_size = 11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA))
```

### Qualitative Analysis

> **[Placeholder — add your findings here]**

*Near-expected counties are analytically interesting as baselines. What makes Brown County "typical"? Are there any tensions between demographic diversity (Green Bay's immigrant and refugee communities) and the structural model's predictions?*

<div class="callout">
<strong>Baseline Case</strong>
[What does "as expected" actually look like on the ground? Document the absence of unusual mobilizing or suppressing factors.]
</div>

---

## Vanderburgh County, IN

```{r vanderburgh-stats}
r <- county_row("Vanderburgh")
```

<div class="county-card middle" style="margin-bottom:1.5rem;">
<div class="card-label">Near-Expected · Indiana</div>
<div class="card-name">Vanderburgh County</div>
<div class="card-stat">Actual turnout: **`r round(r$turnout_rate, 1)`%** &nbsp;|&nbsp; Predicted: `r round(r$predicted, 1)`%</div>
<div class="card-residual">Residual: `r round(r$residual, 2)` percentage points</div>
</div>

### Predictor Profile

```{r vanderburgh-bar, fig.cap="County values vs. state average.", fig.height=3.5}
predictor_bar("Vanderburgh") %>%
  ggplot(aes(x = value, y = variable, fill = type)) +
  geom_col(position = "dodge", width = 0.65) +
  scale_fill_manual(values = c("County" = "#888580", "State Avg" = "#d8d4ce")) +
  labs(x = "Percent (%)", y = NULL, fill = NULL) +
  theme_minimal(base_family = "sans", base_size = 11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA))
```

### Qualitative Analysis

> **[Placeholder — add your findings here]**

*Vanderburgh is home to Evansville, Indiana's third-largest city. Consider the urban-rural divide within the county, Indiana's registration restrictions, and any local competitive dynamics.*

<div class="callout">
<strong>Baseline Case</strong>
[What does "as expected" look like here? Note any notable features even if residual is near zero.]
</div>

---

# Under-Performers {.tabset}

## Clinton County

```{r clinton-stats}
r <- county_row("Clinton")
```

<div class="county-card under" style="margin-bottom:1.5rem;">
<div class="card-label">Under-Performer · `r r$state_name`</div>
<div class="card-name">Clinton County</div>
<div class="card-stat">Actual turnout: **`r round(r$turnout_rate, 1)`%** &nbsp;|&nbsp; Predicted: `r round(r$predicted, 1)`%</div>
<div class="card-residual">Residual: `r round(r$residual, 2)` percentage points</div>
</div>

### Predictor Profile

```{r clinton-bar, fig.cap="County values vs. state average.", fig.height=3.5}
predictor_bar("Clinton") %>%
  ggplot(aes(x = value, y = variable, fill = type)) +
  geom_col(position = "dodge", width = 0.65) +
  scale_fill_manual(values = c("County" = "#c0392b", "State Avg" = "#d8d4ce")) +
  labs(x = "Percent (%)", y = NULL, fill = NULL) +
  theme_minimal(base_family = "sans", base_size = 11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA))
```

### Qualitative Analysis

> **[Placeholder — add your findings here]**

*Suggested dimensions:*

- **Structural barriers:** Voter ID laws, registration restrictions, polling place consolidation?
- **Civic disengagement:** Low competitive elections, weak local party infrastructure, low trust in institutions?
- **Population churn:** High transience (college town, military base, agricultural seasonal workers) that disrupts registration rolls?

<div class="callout warning">
<strong>Emerging Finding</strong>
[Replace with your key takeaway. What explains the gap between structural conditions and actual turnout?]
</div>

---

## DeKalb County

```{r dekalb-stats}
r <- county_row("DeKalb")
```

<div class="county-card under" style="margin-bottom:1.5rem;">
<div class="card-label">Under-Performer · `r r$state_name`</div>
<div class="card-name">DeKalb County</div>
<div class="card-stat">Actual turnout: **`r round(r$turnout_rate, 1)`%** &nbsp;|&nbsp; Predicted: `r round(r$predicted, 1)`%</div>
<div class="card-residual">Residual: `r round(r$residual, 2)` percentage points</div>
</div>

### Predictor Profile

```{r dekalb-bar, fig.cap="County values vs. state average.", fig.height=3.5}
predictor_bar("DeKalb") %>%
  ggplot(aes(x = value, y = variable, fill = type)) +
  geom_col(position = "dodge", width = 0.65) +
  scale_fill_manual(values = c("County" = "#c0392b", "State Avg" = "#d8d4ce")) +
  labs(x = "Percent (%)", y = NULL, fill = NULL) +
  theme_minimal(base_family = "sans", base_size = 11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA))
```

### Qualitative Analysis

> **[Placeholder — add your findings here]**

*Note: If this is DeKalb County, IL (home of NIU), student population effects on CVAP and registration may be particularly relevant. DeKalb, IN would have a different profile — confirm which state before writing up.*

<div class="callout warning">
<strong>Emerging Finding</strong>
[Replace with your key takeaway.]
</div>

---

<span class="section-label">Cross-County Patterns</span>

## Comparing Across Cases

```{r cross-county, fig.cap="All six case-study counties plotted across the five model predictors, with state average benchmarks. Use this to identify cross-cutting patterns.", fig.height=5}
case_counties <- df_pct %>%
  filter(county %in% c("Montmorency", "Trempealeau",
                        "Brown", "Vanderburgh",
                        "Clinton", "DeKalb")) %>%
  mutate(tier = case_when(
    county %in% c("Montmorency", "Trempealeau") ~ "Over",
    county %in% c("Brown", "Vanderburgh")        ~ "Expected",
    TRUE                                          ~ "Under"
  ))

vars <- c("pct_bachelors", "pct_households_internet",
          "pct_limited_english", "pct_disability_total", "pct_poverty")
labels <- c("Bachelor's+", "Internet\nAccess", "Limited\nEnglish",
            "Disability", "Poverty")

case_long <- case_counties %>%
  select(county, tier, all_of(vars)) %>%
  pivot_longer(all_of(vars), names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = vars, labels = labels))

ggplot(case_long, aes(x = variable, y = value, color = tier, group = county)) +
  geom_line(alpha = 0.4, linewidth = 0.7) +
  geom_point(size = 3, alpha = 0.9) +
  ggrepel::geom_text_repel(
    data = case_long %>% filter(variable == "Poverty"),
    aes(label = county), size = 3, family = "sans",
    nudge_x = 0.3, show.legend = FALSE
  ) +
  scale_color_manual(
    values = c("Over" = "#2e7d52", "Expected" = "#888580", "Under" = "#c0392b"),
    name = "Performance Tier"
  ) +
  labs(x = NULL, y = "Percent (%)") +
  theme_minimal(base_family = "sans", base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = NA))
```

*Use this parallel coordinates–style view to identify predictors where over- and under-performers diverge most sharply. Patterns here can motivate hypotheses for qualitative investigation.*

---

<div class="site-footer">
Case county selection based on model residuals. Qualitative analysis draws on [sources to be listed]. Visualizations produced in R with ggplot2.
</div>
