---
title: "Explore County-Level Turnout"
output:
  html_document:
    theme: flatly
    css: custom.css
    highlight: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(tigris_use_cache = TRUE)

library(tidyverse)
library(leaflet)
library(htmlwidgets)
library(htmltools)
library(sf)
library(tigris)

# ── Canonical dataset ───────────────────────────────────────────────────────
canonical_path <- "data/df_pct_with_predictions.rds"
if (!file.exists(canonical_path)) {
  stop("Canonical predictions not found. Run update_data_new.R first.")
}
df_pct <- readRDS(canonical_path)

# Mark case study counties and their performance tier
df_pct <- df_pct %>%
  mutate(
    case_tier = case_when(
      county == "Jasper"      & state_name == "Illinois"   ~ "over",
      county == "Martin"      & state_name == "Indiana"    ~ "over",
      county == "Rock Island" & state_name == "Illinois"   ~ "middle",
      county == "Rusk"        & state_name == "Wisconsin"  ~ "middle",
      county == "Lawrence"    & state_name == "Illinois"   ~ "under",
      county == "Gogebic"     & state_name == "Michigan"   ~ "under",
      TRUE ~ NA_character_
    ),
    is_case        = !is.na(case_tier),
    turnout_pct    = turnout_rate * 100,
    predicted4_pct = predicted4  * 100,
    residual4_pp   = residual4   * 100
  )

# ── County shapefiles ───────────────────────────────────────────────────────
midwest_sf <- counties(
  state        = c("IL", "IN", "OH", "WI", "MI"),
  cb           = TRUE,
  resolution   = "20m",
  year         = 2023,
  progress_bar = FALSE
) %>% st_transform(crs = 4326)

# Join shapefiles with model data
map_sf <- midwest_sf %>%
  mutate(geoid = GEOID) %>%
  left_join(df_pct, by = "geoid")

# Subset case study counties for border highlight layer
case_sf <- map_sf %>% filter(!is.na(case_tier))

# ── Color palettes ──────────────────────────────────────────────────────────
pal_turnout <- colorNumeric(
  palette  = c("#e8f5f2", "#bde5dc", "#7dc8b8", "#3da898", "#2d7f73"),
  domain   = map_sf$turnout_pct,
  na.color = "#dde1e0"
)
pal_predicted <- colorNumeric(
  palette  = c("#e8f5f2", "#bde5dc", "#7dc8b8", "#3da898", "#2d7f73"),
  domain   = map_sf$predicted4_pct,
  na.color = "#dde1e0"
)
# Diverging palette centered near zero: slate (under) → neutral → teal (over)
pal_residual <- colorBin(
  palette  = c("#5a6b67", "#8fa39f", "#c0cfcc", "#f0f2f1",
                "#c5e1db", "#7dc8b8", "#2d7f73"),
  bins     = c(-12, -8, -5, -2, 2, 5, 8, 12),
  na.color = "#dde1e0"
)

# ── Popup HTML builder ──────────────────────────────────────────────────────
build_popup <- function(county, state_name, case_tier,
                        turnout_pct, predicted4_pct, residual4_pp,
                        pct_bachelors, pct_poverty,
                        pct_households_internet, pct_limited_english) {

  badge <- if (!is.na(case_tier)) {
    col <- switch(case_tier,
      "over"   = "#2d7f73",
      "under"  = "#6f7f7b",
      "middle" = "#a8aaa6"
    )
    txt <- switch(case_tier,
      "over"   = "Case Study · Over-performer",
      "under"  = "Case Study · Under-performer",
      "middle" = "Case Study · Near-expected"
    )
    paste0(
      '<div style="font:600 0.68rem/1 \'IBM Plex Sans\',sans-serif;',
      'letter-spacing:.1em;text-transform:uppercase;color:', col,
      ';margin-bottom:.55rem;">', txt, '</div>'
    )
  } else ""

  res_col  <- if (!is.na(residual4_pp) && residual4_pp >= 0) "#2d7f73" else "#6f7f7b"
  res_sign <- if (!is.na(residual4_pp) && residual4_pp >= 0) "+" else ""
  res_str  <- if (!is.na(residual4_pp)) paste0(res_sign, sprintf("%.1f", residual4_pp), " pp") else "—"
  t_str    <- if (!is.na(turnout_pct))    paste0(sprintf("%.1f", turnout_pct), "%")    else "—"
  p_str    <- if (!is.na(predicted4_pct)) paste0(sprintf("%.1f", predicted4_pct), "%") else "—"

  fmt <- function(x, suffix = "%") if (!is.na(x)) paste0(sprintf("%.1f", x), suffix) else "—"

  paste0(
    '<div style="font-family:\'IBM Plex Sans\',sans-serif;',
    'min-width:215px;max-width:260px;line-height:1.45;padding:2px 1px;">',
    badge,
    '<div style="font-family:\'IBM Plex Serif\',Georgia,serif;',
    'font-size:1.05rem;font-weight:600;color:#1f2326;margin-bottom:.12rem;">',
    county, ' County</div>',
    '<div style="font-size:.82rem;color:#8a908d;margin-bottom:.8rem;">', state_name, '</div>',
    # Turnout boxes
    '<div style="display:grid;grid-template-columns:1fr 1fr;gap:.4rem;margin-bottom:.55rem;">',
      '<div style="background:#f7f8f8;padding:.45rem .5rem;border-radius:2px;text-align:center;">',
        '<div style="font-size:.68rem;text-transform:uppercase;letter-spacing:.08em;',
        'color:#8a908d;margin-bottom:.15rem;">Actual</div>',
        '<div style="font-family:\'IBM Plex Serif\',serif;font-size:1.1rem;',
        'font-weight:600;color:#1f2326;">', t_str, '</div>',
      '</div>',
      '<div style="background:#f7f8f8;padding:.45rem .5rem;border-radius:2px;text-align:center;">',
        '<div style="font-size:.68rem;text-transform:uppercase;letter-spacing:.08em;',
        'color:#8a908d;margin-bottom:.15rem;">Model Pred.</div>',
        '<div style="font-family:\'IBM Plex Serif\',serif;font-size:1.1rem;',
        'font-weight:600;color:#1f2326;">', p_str, '</div>',
      '</div>',
    '</div>',
    # Residual
    '<div style="font-size:.88rem;color:#5f6664;margin-bottom:.75rem;">',
    'vs model: <span style="font-weight:600;color:', res_col, ';">', res_str, '</span></div>',
    # Demographics
    '<div style="border-top:1px solid #e2e5e5;padding-top:.6rem;">',
      '<div style="font-size:.68rem;text-transform:uppercase;letter-spacing:.08em;',
      'color:#8a908d;margin-bottom:.35rem;">Key Characteristics</div>',
      '<div style="font-size:.8rem;color:#5f6664;line-height:1.75;">',
        fmt(pct_bachelors), ' Bachelor\'s+ &nbsp;·&nbsp; ', fmt(pct_poverty), ' Poverty<br>',
        fmt(pct_households_internet), ' Internet &nbsp;·&nbsp; ', fmt(pct_limited_english), ' Ltd. English',
      '</div>',
    '</div>',
    '</div>'
  )
}

# Build popup strings for all counties
popup_data <- map_sf %>%
  st_drop_geometry() %>%
  transmute(county, state_name, case_tier, turnout_pct, predicted4_pct, residual4_pp,
            pct_bachelors, pct_poverty, pct_households_internet, pct_limited_english)

map_sf$popup_html <- purrr::pmap_chr(popup_data, build_popup)

# Hover labels
map_sf <- map_sf %>%
  mutate(
    label_base  = paste0(county, " County, ", state_name),
    label_resid = paste0(
      county, " County, ", state_name,
      ifelse(!is.na(residual4_pp),
             paste0("  (", ifelse(residual4_pp > 0, "+", ""), round(residual4_pp, 1), " pp)"),
             "")
    )
  )

# State label anchor points (lower peninsula only for MI)
state_labs <- data.frame(
  name = c("ILLINOIS", "INDIANA", "OHIO", "WISCONSIN", "MICHIGAN"),
  lng  = c(-89.20,    -86.35,   -82.80,  -89.70,     -84.55),
  lat  = c( 40.00,     40.30,    40.50,   44.40,      43.80)
)
```


All 437 counties across Illinois, Indiana, Ohio, Wisconsin, and Michigan — plotted three ways. Toggle between actual turnout, what the model predicts given each county's demographics, and the gap between them.

<div class="callout">
<strong>How to Use</strong>
Use the layer selector (top right of the map) to switch views. Click any county to see a profile card with its actual and predicted turnout, model residual, and key demographic characteristics. Counties selected as case studies are outlined in black — hover to identify them.
</div>

---

```{r county-map, out.width="100%"}
label_opts <- labelOptions(
  style    = list("font-family" = "'IBM Plex Sans', sans-serif",
                  "font-size"   = "12px",
                  "color"       = "#1f2326"),
  textsize = "12px"
)

poly_opts <- function(group, fill_col, label_col) {
  list(
    group       = group,
    fillColor   = fill_col,
    fillOpacity = 0.82,
    color       = "#ffffff",
    weight      = 0.5,
    opacity     = 0.8,
    popup       = htmltools::HTML(map_sf$popup_html),
    label       = label_col,
    labelOptions = label_opts,
    highlightOptions = highlightOptions(
      weight      = 2,
      color       = "#1f2326",
      fillOpacity = 0.92,
      bringToFront = FALSE
    )
  )
}

leaflet(map_sf, height = 580,
        options = leafletOptions(zoomControl = TRUE, scrollWheelZoom = "center")) %>%
  setView(lng = -87.2, lat = 43.1, zoom = 6) %>%

  # Custom pane so state labels + case borders sit above choropleth fill
  addMapPane("top_pane", zIndex = 450) %>%

  # Base tiles — no labels (state names are added manually above polygons)
  addProviderTiles(providers$CartoDB.PositronNoLabels,
                   options = tileOptions(opacity = 0.80)) %>%

  # ── Layer 1: Actual Turnout ──────────────────────────────────────────────
  addPolygons(
    group        = "Actual Turnout",
    fillColor    = ~pal_turnout(turnout_pct),
    fillOpacity  = 0.82,
    color        = "#ffffff",
    weight       = 0.5,
    opacity      = 0.8,
    popup        = ~popup_html,
    label        = ~label_base,
    labelOptions = label_opts,
    highlightOptions = highlightOptions(
      weight = 2, color = "#1f2326", fillOpacity = 0.93, bringToFront = FALSE)
  ) %>%

  # ── Layer 2: Predicted Turnout ───────────────────────────────────────────
  addPolygons(
    group        = "Predicted Turnout",
    fillColor    = ~pal_predicted(predicted4_pct),
    fillOpacity  = 0.82,
    color        = "#ffffff",
    weight       = 0.5,
    opacity      = 0.8,
    popup        = ~popup_html,
    label        = ~label_base,
    labelOptions = label_opts,
    highlightOptions = highlightOptions(
      weight = 2, color = "#1f2326", fillOpacity = 0.93, bringToFront = FALSE)
  ) %>%

  # ── Layer 3: Model Residual ──────────────────────────────────────────────
  addPolygons(
    group        = "Model Residual",
    fillColor    = ~pal_residual(residual4_pp),
    fillOpacity  = 0.82,
    color        = "#ffffff",
    weight       = 0.5,
    opacity      = 0.8,
    popup        = ~popup_html,
    label        = ~label_resid,
    labelOptions = label_opts,
    highlightOptions = highlightOptions(
      weight = 2, color = "#1f2326", fillOpacity = 0.93, bringToFront = FALSE)
  ) %>%

  # ── Case study county borders — gray, above fill ─────────────────────────
  addPolygons(
    data        = case_sf,
    fillColor   = "transparent",
    fillOpacity = 0,
    color       = "#888780",
    weight      = 3,
    opacity     = 1,
    options     = pathOptions(pane = "top_pane", interactive = FALSE)
  ) %>%

  # ── State name labels — permanent markers in elevated pane ────────────────
  addCircleMarkers(
    data        = state_labs,
    lng         = ~lng,
    lat         = ~lat,
    radius      = 0,
    stroke      = FALSE,
    fillOpacity = 0,
    label       = ~name,
    labelOptions = labelOptions(
      noHide    = TRUE,
      direction = "center",
      textOnly  = TRUE,
      style     = list(
        "font-family"    = "'IBM Plex Sans', sans-serif",
        "font-size"      = "13px",
        "font-weight"    = "700",
        "letter-spacing" = "0.12em",
        "color"          = "#1f2326",
        "text-transform" = "uppercase",
        "text-shadow"    = paste(
          "0 0 4px #fff,", "0 0 4px #fff,",
          "1px 1px 0 #fff,", "-1px -1px 0 #fff,",
          "1px -1px 0 #fff,", "-1px 1px 0 #fff"
        ),
        "pointer-events" = "none"
      )
    ),
    options = markerOptions(pane = "top_pane", interactive = FALSE)
  ) %>%

  # ── Layer control ────────────────────────────────────────────────────────
  addLayersControl(
    baseGroups = c("Actual Turnout", "Predicted Turnout", "Model Residual"),
    options    = layersControlOptions(collapsed = FALSE)
  ) %>%

  # ── Legends (one per layer; JS toggles visibility) ───────────────────────
  addLegend(
    position  = "bottomright",
    pal       = pal_turnout,
    values    = map_sf$turnout_pct,
    title     = "Turnout Rate (%)",
    opacity   = 0.85,
    labFormat = labelFormat(suffix = "%")
  ) %>%
  addLegend(
    position  = "bottomright",
    pal       = pal_predicted,
    values    = map_sf$predicted4_pct,
    title     = "Predicted Turnout (%)",
    opacity   = 0.85,
    labFormat = labelFormat(suffix = "%")
  ) %>%
  addLegend(
    position  = "bottomright",
    pal       = pal_residual,
    values    = map_sf$residual4_pp,
    title     = "Residual (pp)",
    opacity   = 0.85,
    labFormat = labelFormat(suffix = " pp")
  ) %>%

  hideGroup("Predicted Turnout") %>%
  hideGroup("Model Residual") %>%

  # ── JS: show/hide legend matching the active layer ───────────────────────
  onRender("
    function(el, x) {
      var map = this;
      var legends = el.querySelectorAll('.info.legend');
      if (legends.length > 1) legends[1].style.display = 'none';
      if (legends.length > 2) legends[2].style.display = 'none';
      var nameToIdx = {
        'Actual Turnout':    0,
        'Predicted Turnout': 1,
        'Model Residual':    2
      };
      map.on('baselayerchange', function(e) {
        for (var i = 0; i < legends.length; i++) {
          legends[i].style.display = 'none';
        }
        var idx = nameToIdx[e.name];
        if (idx !== undefined && legends[idx]) {
          legends[idx].style.display = 'block';
        }
      });
    }
  ")
```

<p style="font-family:'IBM Plex Sans',sans-serif;font-size:0.82rem;color:#8a908d;margin-top:0.5rem;">
Counties outlined in gray are the six case studies. Data: ACS 2020 5-Year Estimates; State Election Commissions (IL, IN, OH, WI, MI); CVAP Special Tabulation. Predictions from Model 4 (12 predictors + state fixed effects).
</p>

---

- **[Background →](descriptive.html)** Descriptive findings: state choropleths and demographic gap analysis
- **[Methodology →](model.html)** Full model specification, results, and residual distribution
- **[Case Studies →](case-studies.html)** Qualitative analysis of six selected counties
- **[Synthesis →](synthesis.html)** Cross-case patterns and implications

---

<div class="site-footer">
Interactive map built with R and <a href="https://rstudio.github.io/leaflet/">Leaflet for R</a>. Choropleth boundaries from the U.S. Census Bureau (TIGER/Line, 2023). Full analysis source on <a href="https://github.com/Smlfberry/midwestturnout">GitHub</a>.
</div>
