---
title: ""
output:
  html_document:
    theme: flatly
    css: custom.css
    highlight: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(tigris_use_cache = TRUE)
try(library(sf), silent = TRUE)
library(tigris)
library(tidycensus)
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggrepel)
source("plot-fonts.R")

canonical_path <- "data/df_pct_with_predictions.rds"
if (!file.exists(canonical_path)) {
  stop("Canonical predictions not found. Run capstoneanalysis.Rmd first.")
}
df_pct <- readRDS(canonical_path)
```

<div class="site-hero">
<div class="hero-grid hero-grid-text-only">
<div class="hero-text">

# What Demographics Can — and Can't — Explain

<p class="dek">A predictive model of county-level turnout across five Midwestern states in the 2024 presidential election identifies counties that vote significantly above or below their demographic expectations. Six cases anchor a mixed-methods investigation into what the data alone cannot capture.</p>

</div>
</div>
</div>

<div class="viz-section">

```{r hero-map, fig.width=11, fig.height=7.5}
options(tigris_use_cache = TRUE)

# County shapefiles for the 5 Midwest states (cb = simplified boundaries)
midwest_counties <- tigris::counties(
  state = c("IL", "IN", "OH", "WI", "MI"),
  cb = TRUE,
  year = 2023,
  progress_bar = FALSE
) |> sf::st_transform(crs = 4326)

# State outlines (dissolve counties to states)
midwest_states <- midwest_counties |>
  dplyr::group_by(STATEFP) |>
  dplyr::summarise(geometry = sf::st_union(geometry), .groups = "drop")

# Join shapefiles with model4 residuals
map_data <- midwest_counties |>
  dplyr::mutate(geoid = GEOID) |>
  dplyr::left_join(df_pct, by = "geoid") |>
  dplyr::mutate(
    residual_pp = residual4 * 100,
    residual_bin = cut(
      residual_pp,
      breaks = c(-Inf, -10, -5, -2, 2, 5, 10, Inf),
      labels = c("≤ −10 pp", "−10 to −5", "−5 to −2",
                 "−2 to +2", "+2 to +5", "+5 to +10", "≥ +10 pp"),
      include.lowest = TRUE,
      ordered_result = TRUE
    )
  )

# Case-study county polygons + label centroids
case_studies <- map_data |>
  dplyr::filter(
    (county == "Jasper"      & state_name == "Illinois")  |
    (county == "Martin"      & state_name == "Indiana")   |
    (county == "Rock Island" & state_name == "Illinois")  |
    (county == "Rusk"        & state_name == "Wisconsin") |
    (county == "Lawrence"    & state_name == "Illinois")  |
    (county == "Gogebic"     & state_name == "Michigan")
  )

case_centroids <- sf::st_drop_geometry(case_studies) |>
  dplyr::mutate(
    centroid_lng = sf::st_coordinates(sf::st_centroid(case_studies$geometry))[, 1],
    centroid_lat = sf::st_coordinates(sf::st_centroid(case_studies$geometry))[, 2]
  )

ggplot() +
  # Counties — colored by Model 4 residual binned into 7 categories
  geom_sf(
    data = map_data,
    aes(fill = residual_bin),
    color = "white",
    linewidth = 0.12
  ) +
  # State outlines — thin, soft warm gray
  geom_sf(
    data = midwest_states,
    fill = NA,
    color = "#A8A095",
    linewidth = 0.3
  ) +
  # Case studies — outlined in warm dark charcoal (neutral, doesn't compete with palette)
  geom_sf(
    data = case_studies,
    fill = NA,
    color = "#2A2724",
    linewidth = 0.55
  ) +
  # County labels via ggrepel
  ggrepel::geom_label_repel(
    data = case_centroids,
    aes(x = centroid_lng, y = centroid_lat, label = county),
    max.overlaps = Inf,
    size = 3,
    box.padding = 0.5,
    segment.color = "gray50",
    segment.size = 0.3
  ) +
  # Discrete diverging palette: deep plum → cream → deep amber (7 bins)
  scale_fill_manual(
    values = c(
      "≤ −10 pp"   = "#5D2A7C",   # deep plum
      "−10 to −5"  = "#7E4FA0",   # medium plum
      "−5 to −2"   = "#B89DCD",   # light plum
      "−2 to +2"   = "#EDE8DA",   # warm cream-gray (near zero)
      "+2 to +5"   = "#E5B07A",   # light amber
      "+5 to +10"  = "#C77E2C",   # medium amber
      "≥ +10 pp"   = "#A8651B"    # deep amber (deeper than before)
    ),
    name = NULL,
    drop = FALSE,
    na.value = "#EDE8DA",
    guide = guide_legend(
      nrow = 1,
      keywidth  = unit(1.6, "cm"),
      keyheight = unit(0.45, "cm"),
      label.position = "bottom",
      label.theme = element_text(
        family = "plex_sans", size = 9, color = "#5A544C"
      )
    )
  ) +
  coord_sf(crs = 4326) +
  labs(
    title    = "How Each County Compared to Its Prediction",
    subtitle = "All 437 counties across five Midwestern states. Case studies outlined in teal.",
    caption  = paste0(
      "Residuals from Model 4 (full specification with state fixed effects). ",
      "Negative values = turnout below prediction; positive = above prediction."
    )
  ) +
  theme_void(base_family = "plex_sans") +
  theme(
    plot.title = element_text(
      family = "plex_serif", face = "bold",
      size = 22, color = "#1F1F1F",
      margin = margin(b = 6), hjust = 0.5
    ),
    plot.subtitle = element_text(
      family = "plex_serif", face = "italic",
      size = 14, color = "#7A746C",
      margin = margin(b = 14), hjust = 0.5
    ),
    plot.caption = element_text(
      family = "plex_sans",
      size = 10, color = "#8C867E",
      hjust = 0.5, margin = margin(t = 14)
    ),
    legend.position = "bottom",
    legend.text  = element_text(family = "plex_sans", size = 10, color = "#5A544C"),
    legend.title = element_text(family = "plex_sans", size = 10, color = "#5A544C"),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(15, 15, 15, 15)
  )
```
</div>

---

## What This Project Does

This project builds on a descriptive needs assessment conducted for the Chicago Lawyers' Committee for Civil Rights (CLC) Midwest Voting Rights Program as part of their Midwest regional expansion. This analysis included identifying demographic patterns associated with low voter turnout across five Midwestern states, finding consistent gaps by education, race, internet access, and disability status. This project moves from identifying descriptive patterns to building a predictive model that identifies counties performing above or below what their demographic data would suggest, then investigating why through qualitative case analysis.

To do so, this project builds four nested OLS regression models predicting county-level turnout rates in the 2024 general election across Illinois, Indiana, Ohio, Wisconsin, and Michigan. Model 1 uses five structural predictors drawn from the American Community Survey. Model 2 adds racial composition and lower-tail educational attainment, informed by the prior CLC descriptive findings. Model 3 introduces state fixed effects, absorbing state-level institutional context like registration rules, ballot access, and voter ID requirements. Model 4 — the most comprehensive and the one used to identify case studies — adds Hispanic, American Indian / Alaska Native, and age-structure variables. The gap between what Model 4 predicts for a county and its actual turnout — called the residual — identifies counties that perform better or worse than their demographic conditions would suggest.

Counties with the largest positive residuals are 'over-performers': something beyond the demographic data captured in the model is driving participation. Counties with the largest negative residuals are 'under-performers': something is suppressing turnout that the model cannot explain. Counties closest to zero are the 'as-expected' cases.

---

## Above and Below the Line

The scatter below pairs the same Model 4 predictions against actual turnout. Counties on the dashed line voted exactly as their structural conditions predicted; those above voted more, those below voted less. The same six case studies highlighted on the map appear here in teal — over-performers in the upper left, under-performers in the lower right.



```{r residual-scatter, fig.width=11, fig.height=7}
df_hero <- df_pct %>%
  mutate(
    highlight = case_when(
      (county == "Jasper"      & state_name == "Illinois")  ~ TRUE,
      (county == "Martin"      & state_name == "Indiana")   ~ TRUE,
      (county == "Rock Island" & state_name == "Illinois")  ~ TRUE,
      (county == "Rusk"        & state_name == "Wisconsin") ~ TRUE,
      (county == "Lawrence"    & state_name == "Illinois")  ~ TRUE,
      (county == "Gogebic"     & state_name == "Michigan")  ~ TRUE,
      TRUE ~ FALSE
    )
  )

# Pre-compute axis limits for annotation positioning
x_range <- range(df_hero$predicted4, na.rm = TRUE)
y_range <- range(df_hero$turnout_rate, na.rm = TRUE)

ggplot(df_hero, aes(x = predicted4, y = turnout_rate)) +

  # Subtle shaded regions to anchor over/under areas (cool slate-blue above line, warm coral below)
  geom_polygon(
    data = data.frame(
      x = c(0.30, 0.30, 1.00),
      y = c(0.30, 1.00, 1.00)
    ),
    aes(x = x, y = y),
    fill = "#FAECE7", alpha = 0.5, inherit.aes = FALSE
  ) +
  geom_polygon(
    data = data.frame(
      x = c(0.30, 1.00, 1.00),
      y = c(0.30, 0.30, 1.00)
    ),
    aes(x = x, y = y),
    fill = "#EEEDFE", alpha = 0.5, inherit.aes = FALSE
  ) +

  # Reference line (perfect prediction)
  geom_abline(
    slope = 1, intercept = 0,
    linetype = "dashed",
    color = "#8E877D",
    linewidth = 0.7
  ) +

  # Background counties
  geom_point(
    data = filter(df_hero, !highlight),
    color = "#D8D1C7",
    size = 1.55,
    alpha = 0.42
  ) +

  # Case-study counties (teal — consistent with site brand)
  geom_point(
    data = filter(df_hero, highlight),
    color = "#2D7F73",
    size = 4.5
  ) +

  ggrepel::geom_text_repel(
    data = filter(df_hero, highlight),
    aes(label = county),
    family = "plex_serif",
    fontface = "bold",
    size = 3.9,
    color = "#2F6F68",
    show.legend = FALSE,
    box.padding = 0.5,
    point.padding = 0.35,
    min.segment.length = 0,
    segment.color = "#AAA297",
    segment.linewidth = 0.28,
    max.overlaps = 20,
    seed = 42
  ) +

  # Annotation: over-performer region (upper left)
  annotate(
    "text",
    x = 0.40, y = 0.93,
    label = "Over-performers",
    family = "plex_serif", fontface = "italic",
    size = 5, color = "#D85A30",
    hjust = 0
  ) +
  annotate(
    "text",
    x = 0.40, y = 0.905,
    label = "Vote more than expected",
    family = "plex_sans",
    size = 3.4, color = "#5A544C",
    hjust = 0
  ) +

  # Annotation: under-performer region (bottom right)
  annotate(
    "text",
    x = 0.78, y = 0.42,
    label = "Under-performers",
    family = "plex_serif", fontface = "italic",
    size = 5, color = "#7F77DD",
    hjust = 0
  ) +
  annotate(
    "text",
    x = 0.78, y = 0.395,
    label = "Vote less than expected",
    family = "plex_sans",
    size = 3.4, color = "#5A544C",
    hjust = 0
  ) +

  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +

  labs(
    title    = "Actual vs. Predicted County Turnout",
    subtitle = "Counties above the dashed line vote more than expected; below the line, less than expected",
    x        = "Predicted turnout rate",
    y        = "Actual turnout rate",
    caption  = paste0("N = ", nrow(df_hero), " counties across five Midwestern states. Predictions from Model 4.")
  ) +

  theme_minimal(base_family = "plex_sans", base_size = 13) +

  theme(
    plot.title = element_text(
      family = "plex_serif", face = "bold",
      size = 26, color = "#1F1F1F",
      margin = margin(b = 6)
    ),
    plot.subtitle = element_text(
      family = "plex_serif", face = "italic",
      size = 15, color = "#7A746C",
      margin = margin(b = 18)
    ),
    axis.title = element_text(size = 12, color = "#4F4A44"),
    axis.text  = element_text(size = 11, color = "#6B655D"),
    panel.grid.major = element_line(color = "#F0ECE4", linewidth = 0.5),
    panel.grid.minor = element_blank(),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.caption = element_text(
      size = 10, color = "#8C867E",
      hjust = 1, margin = margin(t = 12)
    )
  )
```

---

## Case Selection

Two counties from each tier — over-performers, as-expected, and under-performers — were selected for in-depth qualitative analysis. Selection draws from Model 4 residuals (the comprehensive specification) and pairs one extreme case with one moderate case in each tail, so the analysis holds beyond just the most extreme outliers. The two as-expected counties sit at the median rank with residuals near zero — clean baselines where the model predicts almost exactly right.

<div class="county-grid">

<div class="county-card over">
<div class="card-label">Over-Performer · Extreme</div>
<div class="card-name">Jasper County</div>
<div class="card-state">Illinois</div>
<div class="card-stat">Turnout: 77.7% &nbsp;|&nbsp; Predicted: 66.3%</div>
<div class="card-residual">Residual: +11.4 pp</div>
</div>

<div class="county-card over">
<div class="card-label">Over-Performer · Moderate</div>
<div class="card-name">Martin County</div>
<div class="card-state">Indiana</div>
<div class="card-stat">Turnout: 66.1% &nbsp;|&nbsp; Predicted: 57.4%</div>
<div class="card-residual">Residual: +8.7 pp</div>
</div>

<div class="county-card middle">
<div class="card-label">As-Expected</div>
<div class="card-name">Rock Island County</div>
<div class="card-state">Illinois</div>
<div class="card-stat">Turnout: 61.1% &nbsp;|&nbsp; Predicted: 61.0%</div>
<div class="card-residual">Residual: ≈0.0 pp</div>
</div>

<div class="county-card middle">
<div class="card-label">As-Expected</div>
<div class="card-name">Rusk County</div>
<div class="card-state">Wisconsin</div>
<div class="card-stat">Turnout: 73.6% &nbsp;|&nbsp; Predicted: 73.6%</div>
<div class="card-residual">Residual: ≈+0.1 pp</div>
</div>

<div class="county-card under">
<div class="card-label">Under-Performer · Moderate</div>
<div class="card-name">Lawrence County</div>
<div class="card-state">Illinois</div>
<div class="card-stat">Turnout: 50.9% &nbsp;|&nbsp; Predicted: 57.5%</div>
<div class="card-residual">Residual: −6.6 pp</div>
</div>

<div class="county-card under">
<div class="card-label">Under-Performer · Extreme</div>
<div class="card-name">Gogebic County</div>
<div class="card-state">Michigan</div>
<div class="card-stat">Turnout: 70.3% &nbsp;|&nbsp; Predicted: 79.7%</div>
<div class="card-residual">Residual: −9.4 pp</div>
</div>

</div>

---

## Data Sources

<span class="method-tag">ACS 5-Year</span>
<span class="method-tag">CVAP Denominators</span>
<span class="method-tag">State Election Commissions</span>

Turnout rates are calculated using **Citizen Voting Age Population (CVAP)** denominators from ACS estimates. Demographic predictors are drawn from the **2024 ACS 5-Year estimates**:

| Predictor | ACS Table | Rationale |
|-----------|-----------|-----------|
| % Bachelor's degree or higher | B15003 | Civic engagement, information access |
| % No high school diploma | B15003 | Lower-tail educational disadvantage |
| % Households with internet access | B28002 | Voter registration, information environment |
| % Limited English proficiency | B16004 | Ballot access barriers |
| % With a disability | B18101 | Participation barriers, registration gaps |
| % Below poverty line | B17001 | Resource constraints on participation |
| % Black or African American | B02001 | Racial composition |
| % Hispanic or Latino | B03003 | Ethnic composition, language and access barriers |
| % American Indian / Alaska Native | B02001 | Tribal community context, jurisdictional differences |
| % Age 18–24 | B01001 | Young adult / college-age share, lower turnout propensity |
| % Age 65+ | B01001 | Older population, higher turnout propensity |

The full Model 4 specification additionally includes state fixed effects to adjust for state-level institutional context (registration rules, ballot access, election administration). All ACS data were retrieved using the `tidycensus` R package.

---

## Explore This Project

- **[Background →](descriptive.html)** Descriptive findings: state choropleths and demographic gap analysis
- **[Methodology →](model.html)** Full model specification, results, and residual distribution
- **[Case Studies →](case-studies.html)** Qualitative analysis of six selected counties
- **[Synthesis →](synthesis.html)** Cross-case patterns and implications
- **[Explore →](explore.html)** Interactive county-level data explorer

---

<div class="site-footer">
Data: U.S. Census Bureau American Community Survey (2024 5-Year Estimates); CVAP Special Tabulation; State Election Commissions (IL, IN, OH, WI, MI). Analysis conducted in R. Code available on <a href="https://github.com/Smlfberry/midwestturnout">GitHub</a>.
</div>
