---
title: "Methodology"
output:
  html_document:
    theme: flatly
    css: custom.css
    highlight: pygments
    toc: false
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 150)
library(tidyverse)
library(knitr)
library(kableExtra)
library(broom)
library(ggrepel)
library(plotly)

# ── Canonical artifacts ────────────────────────────────────────────────────
# All predictions and model objects come from the canonical Capstone pipeline:
#   /Users/samberry/Desktop/Capstone/update_data_new.R
# Re-run that script whenever the model spec or input data changes.
canonical_df    <- "data/df_pct_with_predictions.rds"
canonical_m1    <- "data/model1.rds"
canonical_m2    <- "data/model2.rds"
canonical_m3    <- "data/model3.rds"
canonical_m4    <- "data/model4.rds"

required_files <- c(canonical_df, canonical_m1, canonical_m2, canonical_m3, canonical_m4)
missing_files  <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop(
    "Canonical artifacts not found:\n  ",
    paste(missing_files, collapse = "\n  "),
    "\nRun /Users/samberry/Desktop/Capstone/update_data_new.R first."
  )
}

df_pct <- readRDS(canonical_df)
model1 <- readRDS(canonical_m1)
model2 <- readRDS(canonical_m2)
model3 <- readRDS(canonical_m3)
model4 <- readRDS(canonical_m4)

# Case study counties (disambiguate by state — Jasper/Lawrence exist in multiple states)
selected <- df_pct %>%
  filter(
    (county == "Jasper"      & state_name == "Illinois")  |
    (county == "Martin"      & state_name == "Indiana")   |
    (county == "Rock Island" & state_name == "Illinois")  |
    (county == "Rusk"        & state_name == "Wisconsin") |
    (county == "Lawrence"    & state_name == "Illinois")  |
    (county == "Gogebic"     & state_name == "Michigan")
  )

# Palette
pal <- c(
  teal       = "#2D7F73",
  teal_light = "#5C9A91",
  slate      = "#6F7F7B",
  cream      = "#F7F8F8",
  rule       = "#E7E1D8",
  text       = "#2D3030",
  text_mid   = "#6F746F",
  text_light = "#9A948C",
  amber      = "#B8772E",
  plum       = "#6E4A8A"
)

source("plot-fonts.R")

# Tidy coefficient table for Model 4, dropping state fixed effects for readability.
# State FE coefficients appear in the model object but aren't substantively interpretable here.
coef_df <- broom::tidy(model4, conf.int = TRUE) %>%
  filter(term != "(Intercept)",
         !str_detect(term, "factor\\(state_name\\)")) %>%
  mutate(
    term = case_when(
      term == "pct_bachelors"           ~ "% Bachelor's Degree or Higher",
      term == "pct_no_hs"               ~ "% No High School Diploma",
      term == "pct_black"               ~ "% Black or African American",
      term == "pct_hispanic"            ~ "% Hispanic or Latino",
      term == "pct_ai_an"               ~ "% American Indian / Alaska Native",
      term == "pct_age_18_24"           ~ "% Age 18-24",
      term == "pct_age_65_plus"         ~ "% Age 65+",
      term == "pct_households_internet" ~ "% Households with Internet",
      term == "pct_limited_english"     ~ "% Limited English Proficiency",
      term == "pct_disability_total"    ~ "% Population with a Disability",
      term == "pct_poverty"             ~ "% Below Poverty Line",
      TRUE ~ term
    ),
    category = case_when(
      term %in% c("% Bachelor's Degree or Higher",
                  "% No High School Diploma",
                  "% Households with Internet") ~ "Education & Access",
      term %in% c("% Black or African American",
                  "% Hispanic or Latino",
                  "% American Indian / Alaska Native",
                  "% Limited English Proficiency") ~ "Race, Ethnicity & Language",
      term %in% c("% Age 18-24", "% Age 65+") ~ "Age Structure",
      TRUE ~ "Economic & Health Context"
    ),
    category = factor(
      category,
      levels = c("Education & Access", "Race, Ethnicity & Language",
                 "Age Structure", "Economic & Health Context")
    ),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ "†",
      TRUE            ~ ""
    ),
    sig_level = case_when(
      p.value < 0.01 ~ "p < .01",
      p.value < 0.05 ~ "p < .05",
      p.value < 0.10 ~ "p < .10",
      TRUE           ~ "Not significant"
    ),
    sig_level = factor(sig_level,
                       levels = c("p < .01", "p < .05",
                                  "p < .10", "Not significant")),
    evidence = if_else(p.value < 0.05, "Significant association", "Unclear / overlaps zero"),
    evidence = factor(evidence,
                      levels = c("Significant association", "Unclear / overlaps zero")),
    # Direction × significance for color encoding
    sig_direction = case_when(
      p.value < 0.05 & estimate > 0 ~ "sig_pos",
      p.value < 0.05 & estimate < 0 ~ "sig_neg",
      estimate > 0                  ~ "ns_pos",
      TRUE                          ~ "ns_neg"
    ),
    estimate_10pp = estimate * 10 * 100,
    conf.low_10pp = conf.low * 10 * 100,
    conf.high_10pp = conf.high * 10 * 100,
    effect_size = abs(estimate_10pp)
  )

s4 <- summary(model4)

fit_table <- bind_rows(
  glance(model1) %>% mutate(Model = "Model 1: CLC framework (5 vars)"),
  glance(model2) %>% mutate(Model = "Model 2: + no HS + Black"),
  glance(model3) %>% mutate(Model = "Model 3: + state fixed effects"),
  glance(model4) %>% mutate(Model = "Model 4: full (age + Hispanic + AI/AN)")
) %>%
  mutate(
    RMSE = sigma * 100  # report RMSE in percentage points
  ) %>%
  select(Model,
         `R²`        = r.squared,
         `Adj. R²`   = adj.r.squared,
         `RMSE (pp)` = RMSE,
         `N`         = nobs)
```


This analysis uses **Ordinary Least Squares (OLS) regression** to model county-level voter turnout rates as a function of structural predictors from the American Community Survey. The unit of analysis is the county; the outcome variable is the county-level turnout rate calculated using CVAP denominators.

Instead of a single specification, the analysis builds up four nested models, each adding variables to the one prior. This progression shows how fit improves with each addition and confirms that Model 4 — which is most comprehensive — is the appropriate basis for case selection.

```{r fit-progression, fig.width=11, fig.height=4.8, fig.cap="Adjusted R-squared and RMSE across the four nested specifications. Model 3 adds state fixed effects; Model 4 is the preferred specification used for case selection."}
fit_plot_df <- fit_table %>%
  mutate(
    Model_short = factor(
      c("Model 1", "Model 2", "Model 3", "Model 4"),
      levels = c("Model 1", "Model 2", "Model 3", "Model 4")
    ),
    Model_detail = c("CLC framework", "+ no HS + Black",
                     "+ state fixed effects", "+ age + race detail"),
    highlight = Model_short %in% c("Model 3", "Model 4")
  )

ggplot(fit_plot_df, aes(x = Model_short, y = `Adj. R²`)) +
  geom_col(
    aes(fill = highlight),
    width = 0.62, alpha = 0.78,
    color = "white", linewidth = 0.5
  ) +
  geom_text(
    aes(label = scales::number(`Adj. R²`, accuracy = 0.01)),
    family = plot_sans, fontface = "bold",
    size = 3.8, color = "#4F4A44",
    vjust = -0.55
  ) +
  geom_text(
    aes(y = pmax(`Adj. R²` - 0.08, 0.08),
        label = paste0("RMSE: ", scales::number(`RMSE (pp)`, accuracy = 0.1), " pp")),
    family = plot_sans, size = 3.1, color = "white",
    fontface = "bold"
  ) +
  geom_text(
    aes(y = -0.075, label = Model_detail),
    family = plot_sans, size = 3.0, color = "#6F746F",
    lineheight = 0.95
  ) +
  scale_y_continuous(
    limits = c(-0.1, 0.92),
    labels = scales::number_format(accuracy = 0.1),
    expand = expansion(mult = c(0, 0.06))
  ) +
  coord_cartesian(clip = "off") +
  scale_fill_manual(
    values = c(`TRUE` = "#2D7F73", `FALSE` = "#C9C5BE"),
    guide = "none"
  ) +
  labs(
    title = "How Much Each Specification Explains",
    subtitle = "Adjusted R-squared rises sharply once state fixed effects enter the model.",
    x = NULL,
    y = "Adjusted R-squared",
    caption = "Bars show adjusted R-squared. In-bar labels report RMSE in turnout percentage points."
  ) +
  theme_minimal(base_family = plot_sans, base_size = 13) +
  theme(
    plot.title = element_text(
      family = plot_serif, face = "bold",
      size = 22, color = "#1F1F1F",
      margin = margin(b = 6)
    ),
    plot.subtitle = element_text(
      family = plot_serif, face = "italic",
      size = 13, color = "#7A746C",
      margin = margin(b = 14)
    ),
    panel.grid.major.y = element_line(color = "#F1EDE6", linewidth = 0.35),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(
      family = plot_sans, face = "plain",
      size = 11.2, color = "#4F4A44",
      margin = margin(t = 5)
    ),
    axis.text.y = element_text(
      family = plot_sans, face = "plain",
      size = 10, color = "#8A837A"
    ),
    axis.title.y = element_text(
      family = plot_sans, face = "plain",
      size = 10.5, color = "#5A544C",
      margin = margin(r = 14)
    ),
    axis.ticks = element_blank(),
    axis.line.x = element_blank(),
    plot.caption = element_text(
      family = plot_sans, size = 10,
      color = "#8C867E", hjust = 1,
      margin = margin(t = 12)
    ),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(12, 18, 28, 15)
  )
```

**Model 1 — original CLC-derived structural predictors (5 variables):**

$$\text{Turnout}_i = \beta_0 + \beta_1(\text{Bachelor's}_i) + \beta_2(\text{Internet}_i) + \beta_3(\text{LEP}_i) + \beta_4(\text{Disability}_i) + \beta_5(\text{Poverty}_i) + \varepsilon_i$$

**Model 2 — adds lower-tail education and racial composition (7 variables):**
adds % no high school diploma and % Black or African American to Model 1.

**Model 3 — adds state fixed effects:**
adds a categorical indicator for state, allowing the intercept to vary by state and absorbing state-level institutional context (registration rules, ballot access, election administration).

**Model 4 — comprehensive specification:**
adds % Hispanic / Latino, % American Indian / Alaska Native, % age 18–24, and % age 65+ to Model 3. This is the preferred specification and drives case selection.

In each model, $i$ indexes counties and $\varepsilon_i$ is the performance gap — how far each county's actual turnout falls from what its demographics predict — and is the quantity of primary analytic interest.

<div class="callout">
<strong>Why OLS?</strong>
Here, the goal is not prediction accuracy but using residuals to inform case selection. OLS produces easily interpretable residuals — stark outliers are the strongest candidates for qualitative investigation.
</div>

---

## Fit Statistics

The largest fit gain comes from adding state fixed effects (Model 2 → Model 3): adjusted R² rises from roughly 0.54 to 0.81, indicating that state-level context absorbs much of the unexplained variance. Adding age structure and additional race variables (Model 3 → Model 4) yields a further gain to roughly 0.86, certainly smaller but still meaningful. Together, this model progression shows why Model 4 is the preferred specification.

```{r fit-table}

fit_table %>%
  kable(digits = c(0, 3, 3, 2, 0),
        align = c("l","r","r","r","r"),
        caption = "Model fit statistics across the four nested specifications. RMSE expressed in percentage points of turnout.") %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = TRUE, font_size = 13)
```

---

## Visual Summary — Model 4

**Model 4 fit:** R² = `r round(s4$r.squared, 3)`, Adjusted R² = `r round(s4$adj.r.squared, 3)`, RMSE = `r round(sqrt(mean(residuals(model4)^2)) * 100, 2)` pp. The model accounts for approximately `r round(s4$r.squared * 100, 1)`% of the variation in county-level turnout. Each point below represents one predictor; the horizontal bar shows how uncertain that estimate is.

```{r coef-plot, fig.width=11, fig.height=6.8, fig.cap="Model 4 coefficient estimates and 95% confidence intervals. Effects are rescaled to percentage-point turnout change for a 10-point increase in each predictor; state fixed effects are estimated but not displayed."}
coef_plot_df <- coef_df %>%
  mutate(
    term_plot = fct_reorder(term, estimate_10pp),
    label_key = if_else(
      p.value < 0.05,
      scales::number(estimate_10pp, accuracy = 0.1, suffix = " pp"),
      NA_character_
    )
  )

ggplot(coef_plot_df, aes(x = estimate_10pp, y = term_plot)) +
  annotate("rect",
           xmin = -1, xmax = 1,
           ymin = -Inf, ymax = Inf,
           fill = "#F7F8F8", alpha = 0.95) +
  geom_vline(xintercept = 0, color = "#9A948C",
             linewidth = 0.5, linetype = "dashed") +
  geom_point(
    aes(color = evidence),
    size = 4.8, alpha = 0.95
  ) +
  facet_grid(category ~ ., scales = "free_y",
             space = "free_y", switch = "y") +
  scale_x_continuous(
    labels = scales::label_number(accuracy = 1, suffix = " pp"),
    breaks = scales::pretty_breaks(n = 7)
  ) +
  scale_color_manual(
    values = c(
      "Significant association" = "#176D5F",
      "Unclear / overlaps zero" = "#C9C5BE"
    ),
    labels = c(
      "Significant association" = "Significant (p < .05)",
      "Unclear / overlaps zero" = "Not significant"
    ),
    drop = FALSE
  ) +
  labs(
    title    = "Which County Traits Move Turnout?",
    subtitle = "Left of zero: associated with lower turnout.  Right of zero: associated with higher turnout.  Teal = statistically significant.",
    x        = "Estimated change in turnout rate (per 10-pp increase in predictor)",
    y        = NULL,
    color    = NULL,
    caption  = "Shaded band marks effects within ±1 pp of zero. State fixed effects estimated but not shown."
  ) +
  theme_minimal(base_family = plot_sans, base_size = 14) +
  theme(
    plot.title = element_text(
      family = plot_serif, face = "bold",
      size = 23, color = "#1F1F1F",
      margin = margin(b = 6)
    ),
    plot.subtitle = element_text(
      family = plot_serif, face = "italic",
      size = 13, color = "#7A746C",
      margin = margin(b = 18)
    ),
    plot.caption = element_text(
      family = plot_sans, size = 10.5,
      color = "#8C867E", hjust = 1,
      margin = margin(t = 12)
    ),
    panel.grid.major.x = element_line(color = "#EDE8DA", linewidth = 0.45),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.title.x = element_text(
      family = plot_sans, size = 12,
      color = "#4F4A44", margin = margin(t = 8)
    ),
    axis.text.x = element_text(family = plot_sans, size = 11.5, color = "#7A746C"),
    axis.text.y = element_text(family = plot_sans, size = 11.5, color = "#4F4A44"),
    legend.position      = "top",
    legend.justification = "left",
    legend.key.size      = unit(0.5, "cm"),
    legend.spacing.x     = unit(0.5, "cm"),
    legend.text          = element_text(family = plot_sans, size = 10.5, color = "#5A544C"),
    plot.title.position   = "plot",
    plot.caption.position = "plot",
    strip.placement   = "outside",
    strip.background  = element_blank(),
    strip.text.y.left = element_text(
      family = plot_sans, face = "bold",
      size = 10.5, color = "#176D5F",
      angle = 0, hjust = 1,
      margin = margin(r = 10)
    ),
    panel.spacing.y  = unit(0.85, "lines"),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin      = margin(15, 20, 15, 15)
  )
```

---

## Residual Distribution — Model 4

The distribution of performance gaps — how far each county's actual turnout falls from its demographic prediction — shows whether the model systematically over- or under-predicts for any subset of counties. A well-calibrated model produces gaps that are approximately normally distributed around zero.

```{r residual-hist, fig.cap="Distribution of Model 4 performance gaps (actual minus predicted turnout, pp). Click the legend to show case-study counties."}
case_pts <- df_pct %>%
  filter(
    (county == "Jasper"      & state_name == "Illinois")  |
    (county == "Martin"      & state_name == "Indiana")   |
    (county == "Rock Island" & state_name == "Illinois")  |
    (county == "Rusk"        & state_name == "Wisconsin") |
    (county == "Lawrence"    & state_name == "Illinois")  |
    (county == "Gogebic"     & state_name == "Michigan")
  ) %>%
  mutate(
    resid_pp    = residual4 * 100,
    y_dot       = 0,
    hover_label = paste0("<b>", county, " County, ", state_name, "</b><br>",
                         "Gap: ", ifelse(resid_pp > 0, "+", ""), round(resid_pp, 1), " pp"),
    # Stagger labels for the two near-zero counties so they don't overlap
    textpos = case_when(
      county == "Rock Island" ~ "top center",
      county == "Rusk"        ~ "bottom center",
      TRUE                    ~ "top center"
    )
  )

sans_font  <- list(family = "IBM Plex Sans, Arial, sans-serif")
serif_font <- list(family = "IBM Plex Serif, Georgia, serif")

plot_ly(height = 440) %>%
  add_histogram(
    data      = df_pct %>% mutate(resid_pp = residual4 * 100),
    x         = ~resid_pp,
    xbins     = list(start = -12, end = 12, size = 1),
    marker    = list(color = "rgba(29,158,117,0.4)", line = list(color = "white", width = 0.5)),
    name      = "All counties",
    showlegend = FALSE,
    hovertemplate = "%{y} counties<extra></extra>"
  ) %>%
  add_trace(
    data = case_pts,
    x    = ~resid_pp,
    y    = ~y_dot,
    type = "scatter",
    mode = "markers+text",
    marker = list(size = 16, color = "#176D5F",
                  line = list(color = "white", width = 2.5)),
    text          = ~county,
    textposition  = ~textpos,
    textfont      = modifyList(sans_font, list(size = 13, color = "#176D5F")),
    name          = "&#9654; Click to view case study counties",
    visible       = "legendonly",
    hovertext     = ~hover_label,
    hovertemplate = "%{hovertext}<extra></extra>"
  ) %>%
  layout(
    title = list(
      text  = "<b>Most Counties Land Close to Their Predicted Turnout</b>",
      font  = modifyList(serif_font, list(size = 20, color = "#1F1F1F")),
      x = 0.01, xanchor = "left", pad = list(b = 8)
    ),
    xaxis = list(
      title     = list(text = "Performance gap (pp — actual minus predicted)",
                       font = modifyList(sans_font, list(size = 12, color = "#4F4A44"))),
      tickfont  = modifyList(sans_font, list(size = 11, color = "#7A746C")),
      zeroline  = FALSE,
      gridcolor = "#F1EDE6",
      range     = c(-12.5, 12.5)
    ),
    yaxis = list(
      title     = list(text = "Number of counties",
                       font = modifyList(sans_font, list(size = 12, color = "#4F4A44"))),
      tickfont  = modifyList(sans_font, list(size = 11, color = "#7A746C")),
      gridcolor = "#F1EDE6",
      tickvals  = list(0, 20, 40, 60),
      range     = list(-6, 76)
    ),
    shapes = list(
      list(type = "rect", x0 = -1, x1 = 1, y0 = 0, y1 = 76,
           fillcolor = "#F7F8F8", opacity = 0.95,
           line = list(width = 0), layer = "below"),
      list(type = "line", x0 = 0, x1 = 0, y0 = 0, y1 = 76,
           line = list(color = "#8E877D", width = 1, dash = "dash"))
    ),
    showlegend = TRUE,
    legend = list(
      orientation = "h", x = 0, y = 1.1,
      font        = modifyList(sans_font, list(size = 14, color = "#176D5F"))
    ),
    annotations = list(list(
      text      = "Negative = turnout below prediction. Shaded band = within ±1 pp.",
      x = 1, y = -0.14, xref = "paper", yref = "paper",
      showarrow = FALSE, align = "right",
      font = modifyList(sans_font, list(size = 10.5, color = "#8A8F86"))
    )),
    margin       = list(l = 60, r = 20, t = 80, b = 60),
    paper_bgcolor = "white",
    plot_bgcolor  = "white",
    bargap        = 0
  )
```

---

## Coefficient Estimates — Model 4

```{r coef-table}
coef_df %>%
  mutate(
    `10-pt Effect (pp)` = estimate_10pp,
    `p-value` = ifelse(p.value < 0.001, "<0.001", formatC(round(p.value, 3), format = "f", digits = 3))
  ) %>%
  select(Predictor = term, Category = category,
         `10-pt Effect (pp)`, `p-value`) %>%
  kable(digits = c(0, 0, 2, 0),
        align = c("l","l","r","r"),
        caption = "Model 4 OLS coefficients (state fixed effects suppressed for readability). Outcome: county-level turnout rate (CVAP denominator). N = counties across IL, IN, OH, WI, MI.") %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = TRUE, font_size = 13) %>%
  column_spec(1, bold = FALSE) %>%
  footnote(general = "The 10-point effect column shows the expected change in county turnout rate for a 10-point increase in each predictor. For example, a −6.6 pp effect for % No High School Diploma means that counties with 10 percentage points more residents without a high school diploma tend to have turnout 6.6 points lower, holding all other predictors constant.",
           general_title = "Note: ")
```

---

## Selected Counties: Full Statistics — Model 4

```{r selected-table}
selected_sorted <- selected %>% arrange(desc(residual4))

resid_vals <- round(selected_sorted$residual4 * 100, 1)
resid_colors <- case_when(
  abs(resid_vals) <= 0.15 ~ "#888780",
  resid_vals > 0          ~ "#D85A30",
  TRUE                    ~ "#7F77DD"
)

selected_sorted %>%
  mutate(
    Tier           = case_when(
      residual4 >  0.01 ~ "Over-Performer",
      residual4 < -0.01 ~ "Under-Performer",
      TRUE              ~ "As-Expected"
    ),
    `Actual (%)`    = round(turnout_rate * 100, 1),
    `Predicted (%)` = round(predicted4   * 100, 1),
    `Residual (pp)` = resid_vals,
    `Rank (m4)`     = as.integer(rank_m4)
  ) %>%
  select(County = county, State = state_name, Tier,
         `Actual (%)`, `Predicted (%)`, `Residual (pp)`, `Rank (m4)`) %>%
  kable(align = c("l","l","l","r","r","r","r"),
        caption = "The six case-study counties with Model 4 statistics. Residuals in percentage points. Rank out of 437 counties (1 = most underperforming, 437 = most overperforming).") %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = TRUE, font_size = 13) %>%
  column_spec(6, color = resid_colors, bold = TRUE) %>%
  row_spec(which(selected_sorted$residual4 >  0.01), background = "#FAECE7") %>%
  row_spec(which(selected_sorted$residual4 < -0.01), background = "#EEEDFE")
```

The selection pairs **one extreme + one moderate** county in each tail (over- and under-performers) so the qualitative analysis can analyze beyond simply the most extreme outliers. The two as-expected counties, Rock Island (IL) and Clark (IN), have residuals near zero. This provides a clean baseline where Model 4 predicts the turnout almost exactly.

---

<div class="callout">
<strong>Limitations</strong>
This model is cross-sectional and observational. As such, it is important to note that the coefficients should not be interpreted as causal effects.
</div>

---

- **[Background →](descriptive.html)** Descriptive findings: state choropleths and demographic gap analysis
- **[Case Studies →](case-studies.html)** Qualitative analysis of six selected counties
- **[Synthesis →](synthesis.html)** Cross-case patterns and implications
- **[Explore →](explore.html)** Interactive county-level data explorer

---
